---
title: "Data Wrangling - Part II"
subtitle: "StatPREP R Workshops"
author: "Elisa Gonzalez"
output: html_document
---

```{r setup, include=FALSE}
# Don't forget to load your libraries
library(tidyverse)
library(ggplot2movies)
library(ggformula)
```

```{r}
# This loads the data from the ggplot2movies package
data(movies)
```

# Deriving information

So far, you have learned "verbs" for isolating data, `select`, `filter`, `arrange`. Now, we will learn verbs for deriving information, `summarize`, and `mutate`. There is an additional verb, `group_by`, which is useful in conjunction with these verbs, but not as much on its own. 

## summarize

Summarize is a many-to-one or many-to-few operation. It summarizes the dataset down to fewer numbers. 

Let's say we want to know the average budget of the movies in the `movies` dataset. The way to compute that is with summarize,

```{r}
movies %>%
  summarize(mean(budget))
```

Unfortunately, there are many missing (or Not Available) values in `budget`, so we need to remove those in our pipeline.

```{r}
movies %>%
  drop_na(budget) %>%
  summarize(mean(budget))
```

Notice that this pipeline does not change the `movies` dataset in our environment. The NA values are still present in the data, and the entire dataset is just how it started.  We just `summarize`d the data down to a single number in our Console or RMarkdown document. 

If we want, we can add additional statistics to our `summarize` command,

```{r}
movies %>%
  drop_na(budget) %>%
  summarize(mean(budget), max(budget))
```

We can also name the statistics we compute, so we could refer to them later.

```{r}
movies %>%
  drop_na(budget) %>%
  summarize(avg_budget = mean(budget), max_budget = max(budget))
```

### Your turn

Now, you try it! 

- What year was the earliest movie in the dataset released? 
- What is the average length of movies?
- What is the standard deviation of length?

You may want to look at the "data transformation with dplyr" cheatsheet to review some possible summary functions. 

#Question 1

```{r eval=TRUE}
movies_lin71<-movies %>%
  summarize(mean(length),sd(length),min(year))
movies_lin71
```
The earliest films on this list date back to 1893. The average runtime of the movies is approximately 82 minutes and 34 seconds, with a standard deviation of around 44 minutes and 35 seconds.    

#Question 2   

## group_by

The `group_by` function groups cases by common values. 

This is useful if want to compute summary statistics by group. For example, we may be interested in whether action movies are longer than non-action movies. 

We already know how to compute the average length of movies overall. What we need to add to our pipeline is a `group_by`

```{r eval=TRUE}
movies_ln84<-movies %>%
  group_by(Action) %>%
  summarize(mean(length),sd(length),min(year))
  # move your summarize code from the last exercise here!
movies_ln84
```

Which movies are longer, action movies or non-action movies? 

On average, action movies tend to be longer, with runtimes of around 100 minutes, while non-action movies have an average runtime of about 81 minutes. The standard deviation further highlights this difference, as action movies show less variation, with most films clustering around the 100-minute mark, while non-action movies exhibit more variability in their runtimes.   

#Question 3  

### Your turn

Investigate the movies classified as Short. Are they actually shorter than other movies? 

- What is the average length? 
- What is the minimum length?

```{r eval=TRUE}
movies_ln99<-movies %>%
  group_by(Short) %>%
  summarize(mean(length),min(length))
movies_ln99
```

Short films have an average runtime of around 14 minutes, whereas non-short films typically average about 95 minutes. Interestingly, both short and non-short films share the same minimum runtime of 1 minute, representing the shortest possible length for a movie.   

## mutate

The `mutate` function creates new variables (columns) in the dataset. Typically, we use it to make new variables from old ones. 

For example, we may be interested in the length of movies in hours, rather than in minutes. 

```{r}
movies <- movies %>%
  mutate(length_hrs = length / 60)
```

Let's read this code carefully. 

- what does the `<-` do?
- why are we writing `/60`?
- what does the `=` do?
- why did we use the name `length_hrs`?

### Your turn

#Question 4  

The variables `r1`, `r2`, ..., `r10` give you the percent of people who rated that movie a 1, 2, ..., 10. We would like to know the percent of people who rated the movie a 4 or less (i.e., who rated it a 1, 2, 3, or 4). Create a new variable called `per_low` that computes that number.

```{r eval=TRUE}
movies_ln128 <- movies %>%
  mutate(per_low = r1+r2+r3+r4)
movies_ln128
```

#Question 5 

## Putting it together

Now that we have a bunch of wrangling skills, let's put them together. 

- pull out the movies that had a high percent of low ratings (Hint: what command will you use to do this?)
- pull out movies that had a high percent of low ratings *and* more than 5 votes. 
- how many of these bad movies were there?

To accomplish that last task, we need a specialty function, `n`. Here's how it works

The "low rating" is defined in the previous question using per_low, which identifies movies with low ratings. We will then filter this data to include only movies with a per_low rating and more than 5 votes. This will give us a dataset of movies that are considered low-rated and have received more than 5 votes. Our next step is to calculate the total number of movies that meet these criteria.

```{r}
movies_ln144<-movies_ln128 %>%
  filter(per_low > 50 & votes > 5)%>%
  select(per_low,votes,title) %>%
  summarize(Bdmvs = n())
movies_ln144
```

A total of 4,753 movies have a high percentage of low ratings and receive more than 5 votes.   

#Side Questions 

- what is the average length of movies in the dataset, in hours? 
- can you find a second way to compute the average length, by flipping the order of commands (i.e., `summarize` first and then `mutate`)?

```{r}
average_length_hours <- movies %>%
summarize(average_length = mean(length)) %>%
mutate(average_length_hours_summarize = average_length / 60)

average_length_hours_flipped <- movies %>%
mutate(length_hours = length / 60) %>%
summarize(average_length_hours_mutate = mean(length_hours))

average_length_hours
average_length_hours_flipped

``` 

Flipping the order of the mutate and summarize operations was possible, as demonstrated by the new data sets, average_length_hours and average_length_hours_flipped. Using mutate first provides a more straightforward solution, as it directly gives us the average length in hours when needed. On the other hand, applying summarize first and then mutate shows the step-by-step process that led to the data, which can be useful for tracking progression.   


